prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-21 inst/MyArticle/paper.Rmd
changes to conclusion.

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats and enables the use of the methods without additional arguments. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy data. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (Supplement Figure SX).

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method to convert the `LFQData` class into a `r BiocStyle::Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code, where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`s functionality (see Supplement Figure S[X]).

@@

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:FDRfdp) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:FDRfdp) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (supplement). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see supplement). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:FDRfdp) Panel C) are comparable for all three methods.

>>>>

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats and enables the use of the methods without additional arguments. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy data. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method to convert the `LFQData` class into a `r BiocStyle::Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code, where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`s functionality (see Supplementary Figure S[X]).

@@

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:FDRfdp) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:FDRfdp) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (see Supplement, Benchmark CPTAC). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see Supplement, Testing ROC curves). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:FDRfdp) Panel C) are comparable for all three methods.
